---
title: "LB-SPR tutorial"
output: rmarkdown::html_vignette
vignette: >
    %\VignetteIndexEntry{LB-SPR tutorial}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
```{r, setup}
library(magrittr) # required for this vignette to work
```
# 1. Install `lbmstoolbox`

To run this vignette, install `lbmstoolbox` in one of R's library paths. Install it system-wide with `devtools::install_github()` or locally with `renv::install()`. When running the source interactively, use the `vignettes` directory as the working directory so that the bundled data file can be found.

# 2. Read sampling and biological data

The bundled example contains catch and biological data. Its first element, `catch_data`, holds the catch length composition in two forms: a **long** data frame with year–length observations in rows, and a **wide** data frame with length classes in rows and years in columns. Both use `MeanLength` for the length-class column.

The second element contains the biological parameters for the evaluated species.

```{r read_sampling_data}
spp_sampling_details <- readRDS("data/spp_test_data.rds")
```
## 2.1 Long data frame
```{r long_dataframe_showcase}
catch_data <- spp_sampling_details$catch_data
catch_data$long %>%
  head(6)
```
## 2.2 Wide data frame
```{r wide_dataframe_showcase}
catch_data$wide %>%
  head(6)
```
## 2.3 Biological parameters
```{r biological_parameters}
spp_sampling_details$bio_params
```
# 3. Run LB-SPR

## 3.1 Build biological parameters for LB-SPR

See the [LBSPR documentation](https://adrianhordyk.github.io/LBSPR/articles/LBSPR.html) for parameter definitions. `fecb` is the exponent in the size–fecundity relationship. LB-SPR defaults to `3`; here, it is set to the supplied exponent from the weight–length relationship.

The example biological-parameter list contains fields for both LB-SPR and LIME. Remove the LIME-specific fields and derive LB-SPR's `mk` parameter.

```{r biological_parameters_lbspr}
bio_params <- spp_sampling_details$bio_params

# We need to provide MK
bio_params$mk <- round(bio_params$M / bio_params$k, 2)
# ... And get rid of those that not required
bio_params$M <- NULL
bio_params$t0 <- NULL
bio_params$max_age <- NULL
bio_params$rec_variability_mean <- NULL
bio_params$rec_variability_sd <- NULL
bio_params
```
## 3.2 Run LB-SPR
With `simul = TRUE`, the wrapper runs LB-SPR's simulator for each time step and returns:

- Observed catch-at-length
- Observed catch-at-length, scaled to sum to one
- Standardised fished and unfished population-at-length structures
- Standardised estimated catch after applying estimated selectivity to the fished population
- Standardised estimated catch after applying estimated selectivity to the unfished population

```{r run_lbspr}
iim_lbspr <- lbmstoolbox::LbsprLbms$new(
  bio_params,
  catch_data
)
result_lbspr <- iim_lbspr$run(simul = TRUE)
```
## 3.3 Breaking down results
Results have two components: **evaluation**, which contains model estimates, and **simulation**, which contains the additional simulation output. The latter also retains the raw LB-SPR simulation output.

The estimates data frame also includes confidence intervals for `FM` and `SPR`, with `lower_` and `upper_` prefixes.

### 3.3.1 Estimates dataframe
```{r lbspr_estimates}
result_lbspr$evaluation$estimates %>%
  head(3)
```

### 3.3.2 Simulation dataframes
The simulation data frame includes the following columns for each time step:

- `unfished_pop`: length structure of the unfished population
- `fished_pop`: length structure of the fished population
- `unfished_vul`: estimated catch from the unfished population after applying selectivity
- `exp_catch`: estimated catch from the fished population after applying selectivity
- `real_catch`: observed catch in absolute numbers
- `relative_catch`: observed catch scaled to sum to one

```{r lbspr_simulation}
result_lbspr$simulation$model_info %>%
  head(3)
```
# 4. Plot examples

## 4.1 Estimates plot
A quick example on how to plot SPR and its confidence interval is shown below.

```{r lbspr_estimate_plots, fig.width=8}
estimates <- result_lbspr$evaluation$estimates
g <- estimates %>%
  ggplot2::ggplot(ggplot2::aes(x = years, y = SPR)) +
  ggplot2::geom_point() +
  ggplot2::geom_line() +
  ggplot2::geom_ribbon(ggplot2::aes(x = years, ymin = lower_spr, ymax = upper_spr, fill = "maroon2", alpha = 0.1)) +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "none")

g
```

## 4.2 Simulation plot
The first plot compares estimated catch from the fished and unfished populations in 2015. The second compares the corresponding population-at-length structures.

### 4.2.1 Estimated catch

```{r lbspr_estimate_catch_plots, fig.width=8}
catch_2015 <- result_lbspr$simulation$model_info %>%
  dplyr::filter(year == 2015) %>%
  dplyr::select(lengths, unfished_vul, exp_catch) %>%
  tidyr::pivot_longer(!lengths, names_to = "variables", values_to = "values")

g <- catch_2015 %>%
  ggplot2::ggplot(ggplot2::aes(x = lengths, y = values, color = variables)) +
  ggplot2::geom_point() +
  ggplot2::geom_line() +
  ggplot2::xlab("Lengths") +
  ggplot2::ylab("Standardised relative catch") +
  ggplot2::scale_color_manual(
    name = "Catch Type",
    values = c(
      unfished_vul = "seagreen",
      exp_catch = "tomato"
    )
  )

g
```

### 4.2.2 Estimated populations

```{r lbspr_estimate_population_plots, fig.width=8}
population_2015 <- result_lbspr$simulation$model_info %>%
  dplyr::filter(year == 2015) %>%
  dplyr::select(lengths, unfished_pop, fished_pop) %>%
  tidyr::pivot_longer(!lengths, names_to = "variables", values_to = "values")

g <- population_2015 %>%
  ggplot2::ggplot(ggplot2::aes(x = lengths, y = values, color = variables)) +
  ggplot2::geom_point() +
  ggplot2::geom_line() +
  ggplot2::xlab("Lengths") +
  ggplot2::ylab("Standardised relative populations") +
  ggplot2::scale_color_manual(
    name = "Catch Type",
    values = c(
      unfished_pop = "seagreen",
      fished_pop = "tomato"
    )
  )

g
```
