---
title: "LB-SPR tutorial"
output: rmarkdown::html_vignette
vignette: >
    %\VignetteIndexEntry{LB-SPR tutorial}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
```{r, setup}
library(magrittr) # required for this vignette to work
```
# 1. Installing `lbmstoolbox` package
For this vigenette to work the `lbmstoolbox` package  must be installed and exist in whatever R's libpath of yours. To do so you
can install it widely on the system by using `devtools::install_github(..)` or locally by `renv::install`. if ran interactively,
ensure that the working directory is set to the `vignettes` folder - use `setwd` as suits you best.

# 2. Read sampling and biological data
A data sample has been provided to showcase how both algorithm interfaces work. The first element of the list is the catch's
length composition that the model will assess. This structure, in turn, stores two dataframes: i) a **long** one where years and
lengths are laid as rows and ii) a **wide** one where years are set as columns and lengths as rows. Both dataframes contain `MeanLength`,
the name of the column holding the length classes.

The second element is a list with the biological parameters of the species being evaluated, which we will call `spp.demersal`.

```{r read_sampling_data}
spp_sampling_details <- readRDS("data/spp_test_data.rds")
```
## 2.1 Showcase long dataframe
```{r long_dataframe_showcase}
catch_data <- spp_sampling_details$catch_data
catch_data$long %>%
  head(6)
```
## 2.2 Showcase wide dataframe
```{r wide_dataframe_showcase}
catch_data$wide %>%
  head(6)
```
## 2.3 Showcase wide dataframe
```{r biological_parameters}
spp_sampling_details$bio_params
```
# 3. Run LB-SPR

## 3.1 Build biological parameters required by LB-SPR
Please visit [LBSPR documentation](https://adrianhordyk.github.io/LBSPR/articles/LBSPR.html) for further details. `fecb`
defines the relationship between size and fecundity. By default, LB-SPR sets it to 3. We use the provided `b` exponent
obtained from the weight-length relationship.

The biological parameters here provided contains the union of those required by LB-SPR and LIME. Let's strip off LIME's
specific parameters and build LBSPR'S missing ones.

```{r biological_parameters_lbspr}
bio_params <- spp_sampling_details$bio_params

# We need to provide MK
bio_params$mk <- round(bio_params$M / bio_params$M, 2)
# ... And get rid of those that not required
bio_params$M <- NULL
bio_params$t0 <- NULL
bio_params$max_age <- NULL
bio_params$rec_variability_mean <- NULL
bio_params$rec_variability_sd <- NULL
bio_params
```
## 3.2 Run LB-SPR
If LB-SPR is run with the flag `simul` set to TRUE, re-runs the simulation function to provide extra details for each
time step used as follows:

- Observed catch-at-length
- Relative observed catch-at-length (each catch is divided by the maximum)
- Standardised length structure of the underlying fished population
- Standardised length structure of the underlying unfished population
- Standardised estimated catch as a result of applying the estimated selectivity to the estimated fished population
- Standardised estimated catch as a result of applying the estimated selectivity to the estimated unfished population

```{r run_lbspr}
iim_lbspr <- lbmstoolbox::LbsprLbms$new(
  bio_params,
  catch_data
)
result_lbspr <- iim_lbspr$run(simul = TRUE)
```
## 3.3 Breaking down results
Results are provided in two flavors: i) **evaluation** where estimates of the model are held and
ii) **simulation** where the aforementioned extra details can be found. Within this section, also the raw output yielded by
the simulation function are provided, should you be interested in doing more with it.

In addition, the estimate dataframe provides the confidence intervals of `FM` and `SPR`. These columns are prefixed by `lower` and `upper`, respectively.

### 3.3.1 Estimates dataframe
```{r lbspr_estimates}
result_lbspr$evaluation$estimates %>%
  head(3)
```

### 3.3.2 Simulation dataframes
Below an example of the simulation dataframe for one single time step. The relevant columns are:

- `unfished_pop`: length structure of the unfished population.
- `fished_pop`: length structure of the fished population.
- `unfished_vul`: estimated catch yielded by the estimated unfished population and selectivity.
- `expected_catch`: estimated catch yielded by the estimated fished population and selectivity.
- `real_catch`: observed catch in absolute numbers
- `relative_catch`: Relative-to-maximum observed catch
-

```{r lbspr_simulation}
result_lbspr$simulation$model_info %>%
  head(3)
```
# 4. Plot examples

## 4.1 Estimates plot
A quick example on how to plot SPR and its confidence interval is shown below.

```{r lbspr_estimate_plots}
estimates <- result_lbspr$evaluation$estimates
g <- estimates %>%
  ggplot2::ggplot(ggplot2::aes(x = years, y = SPR)) +
  ggplot2::geom_point() +
  ggplot2::geom_line() +
  ggplot2::geom_ribbon(ggplot2::aes(x = years, ymin = lower_spr, ymax = upper_spr, fill = "maroon2", alpha = 0.1)) +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "none")

g
```
## 4.2 Simulation plot
Below two different plots are displayed. The first one shows the estimated catch from both the estimated fished and unfished
length structured for year 2015. The second one plots the estimated fished and unfished populations for the same year.

### 4.2.1 Estimated catch

```{r lbspr_estimate_catch_plots}
catch_2015 <- result_lbspr$simulation$model_info %>%
  dplyr::filter(year == 2015) %>%
  dplyr::select(lengths, unfished_vul, exp_catch) %>%
  tidyr::pivot_longer(!lengths, names_to = "variables", values_to = "values")

g <- catch_2015 %>%
  ggplot2::ggplot(ggplot2::aes(x = lengths, y = values, color = variables)) +
  ggplot2::geom_point() +
  ggplot2::geom_line() +
  ggplot2::xlab("Lengths") +
  ggplot2::ylab("Standardised relative catch") +
  ggplot2::scale_color_manual(
    name = "Catch Type",
    values = c(
      unfished_vul = "seagreen",
      exp_catch = "tomato"
    )
  )

g
```
### 4.2.1 Estimated populations

```{r lbspr_estimate_population_plots}
population_2015 <- result_lbspr$simulation$model_info %>%
  dplyr::filter(year == 2015) %>%
  dplyr::select(lengths, unfished_pop, fished_pop) %>%
  tidyr::pivot_longer(!lengths, names_to = "variables", values_to = "values")

g <- population_2015 %>%
  ggplot2::ggplot(ggplot2::aes(x = lengths, y = values, color = variables)) +
  ggplot2::geom_point() +
  ggplot2::geom_line() +
  ggplot2::xlab("Lengths") +
  ggplot2::ylab("Standardised relative populations") +
  ggplot2::scale_color_manual(
    name = "Catch Type",
    values = c(
      unfished_pop = "seagreen",
      fished_pop = "tomato"
    )
  )

g
```
