---
title: "LIME tutorial"
output: rmarkdown::html_vignette
vignette: >
    %\VignetteIndexEntry{LIME tutorial}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
**Note:** The model-fitting chunks in this vignette are disabled because of a `Matrix` compatibility issue. See the [README](https://d2gex.github.io/lbmstoolbox/) for details.
```{r, setup}
library(magrittr) # required for this vignette to work
```
# 1. Install `lbmstoolbox`

To run this vignette, install `lbmstoolbox` in one of R's library paths. Install it system-wide with `devtools::install_github()` or locally with `renv::install()`. When running the source interactively, use the `vignettes` directory as the working directory so that the bundled data file can be found.

# 2. Read sampling and biological data
See the [LB-SPR tutorial](../articles/lbspr_use_case.html) for details of the required input data.

```{r read_sampling_data}
spp_sampling_details <- readRDS("data/spp_test_data.rds")
catch_data <- spp_sampling_details$catch_data
```

# 3. Run LIME model

## 3.1 Build biological parameters for LIME

See the [LIME documentation](https://github.com/merrillrudd/LIME/blob/master/vignettes/LIME.pdf) for parameter definitions.

The example biological-parameter list contains fields for both LB-SPR and LIME. Remove the LB-SPR-specific field. `rec_variability_mean` and `rec_variability_sd` define the mean and standard deviation of LIME's recruitment-variability prior (`SigRprior`), preventing unrealistically variable recruitment estimates (Rudd and Thorson, 2017). The example values are based on Thorson et al. (2014).

```{r biological_parameters_IME}
bio_params <- spp_sampling_details$bio_params
bio_params$fecb <- NULL
bio_params
```
## 3.2 Build exploitation parameters for LIME

Unlike LB-SPR, LIME requires selectivity priors and a penalty that constrains year-to-year changes in estimated fishing mortality (Rudd and Thorson, 2017). Here, `SL50` is halfway between the smallest observed length class and the modal class. `SL95` is set to `SL50 * 1.15`, following the approach used by Prince et al. (2022) for the maturation ogive. `sigmaF` uses the default value of `0.2` from Rudd and Thorson (2017).

```{r exploitation_parameters_IME}
summary_catch <- catch_data$long %>%
  dplyr::select(MeanLength, catch) %>%
  dplyr::group_by(MeanLength) %>%
  dplyr::summarise(total_catch = sum(catch))
catch_mode <- summary_catch$MeanLength[which.max(summary_catch$total_catch)]
sl50 <- median(min(summary_catch$MeanLength):catch_mode)
sel_params <- list(
  sl50 = sl50,
  sl95 = sl50 * 1.15
)

exploitation_params <- list(
  s50 = sel_params$sl50,
  s95 = sel_params$sl95,
  sigmaF = 0.2
)
exploitation_params
```
## 3.3 Build the remaining LIME context

As an age-structured model, LIME must be told whether selectivity and maturity are supplied by length or age. This list also sets steepness to `h = 0.7`.

```{r extra_details_IME}
extra_lht_details <- list(
  binwidth = 1,
  selex_input = "length",
  maturity_input = "length",
  nseasons = 1,
  nfleets = 1,
  h = 0.7
)
```

## 3.4 Instantiate and run LIME
With `simul = TRUE`, the wrapper returns the following additional output for each time step:

- Observed catch-at-length
- Observed catch-at-length, scaled to sum to one
- Standardised fished and unfished population-at-age structures
- Standardised estimated catch-at-length after applying estimated selectivity to the fished population
- Standardised estimated catch-at-length after applying estimated selectivity to the unfished population

```{r run_lime, eval = FALSE}
# Find SL50 and SL95 prior
iim_lime_logistic <- lbmstoolbox::LimeLbms$new(
  bio_params,
  exploitation_params,
  catch_data
)
lht_context <- iim_lime_logistic$build_lht_context(extra_lht_details)
result_lime <- iim_lime_logistic$run(lht_context, simul = TRUE)
```
## 3.5 Breaking down results
Results have two components: **evaluation**, which contains model estimates, and **simulation**, which contains the additional output. The latter also retains the raw LIME output.

The estimates data frame also includes confidence intervals for `F`, `SPR`, and recruitment (`R`), with `lower_` and `upper_` prefixes.

### 3.5.1 Estimates dataframe
```{r lime_estimates, eval = FALSE}
result_lime$evaluation$estimates %>%
  head(3)
```

### 3.5.2 Simulation dataframes
The simulation output contains separate catch and population data frames. Their relevant columns are:

- `unfished_pop`: age structure of the unfished population
- `fished_pop`: age structure of the fished population
- `unfished_vul`: estimated catch from the unfished population after applying selectivity
- `exp_catch`: estimated catch from the fished population after applying selectivity
- `real_catch`: observed catch in absolute numbers
- `relative_catch`: observed catch scaled to sum to one

Unlike LB-SPR, LIME returns catch and population structures separately because catch is described by length and population by age.

```{r lime_simulation_catch, eval = FALSE}
result_lime$simulation$model_info$catch %>%
  head(3)
```

```{r lime_simulation_population, eval = FALSE}
result_lime$simulation$model_info$population %>%
  head(3)
```

# 4. Plot examples

## 4.1 Estimates plot
A quick example on how to plot SPR and its confidence interval is shown below.

```{r lbspr_estimate_plots, eval = FALSE}
estimates <- result_lime$evaluation$estimates
g <- estimates %>%
  ggplot2::ggplot(ggplot2::aes(x = years, y = SPR)) +
  ggplot2::geom_point() +
  ggplot2::geom_line() +
  ggplot2::geom_ribbon(ggplot2::aes(x = years, ymin = lower_spr, ymax = upper_spr, fill = "maroon2", alpha = 0.1)) +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "none")

g
```
## 4.2 Simulation plot
The first plot shows estimated catch-at-length for the fished population in 2015. The second compares the fished and unfished population-at-age structures for the same year.

### 4.2.1 Estimated catch

```{r lime_estimate_catch_plots, eval = FALSE}
catch_2015 <- result_lime$simulation$model_info$catch %>%
  dplyr::filter(year == 2015) %>%
  dplyr::select(lengths, exp_catch)

g <- catch_2015 %>%
  ggplot2::ggplot(ggplot2::aes(x = lengths, y = exp_catch)) +
  ggplot2::geom_point() +
  ggplot2::geom_line() +
  ggplot2::xlab("Lengths") +
  ggplot2::ylab("Standardised relative catch")

g
```
### 4.2.2 Estimated populations

```{r lime_estimate_population_plots, eval = FALSE}
population_2015 <- result_lime$simulation$model_info$population %>%
  dplyr::filter(year == 2015) %>%
  dplyr::select(age, unfished_pop, fished_pop) %>%
  tidyr::pivot_longer(!age, names_to = "variables", values_to = "values")

g <- population_2015 %>%
  ggplot2::ggplot(ggplot2::aes(x = age, y = values, color = variables)) +
  ggplot2::geom_point() +
  ggplot2::geom_line() +
  ggplot2::xlab("Ages") +
  ggplot2::ylab("Standardised relative populations") +
  ggplot2::scale_color_manual(
    name = "Catch Type",
    values = c(
      unfished_pop = "seagreen",
      fished_pop = "tomato"
    )
  )

g
```
# 5. References

1. Rudd, M.B., Thorson, J.T., 2017. Accounting for Variability and Biases in Data-limited Fisheries Stock Assessment. Canadian Journal of Fisheries and Aquatic Sciences 75, 1019–1035. https://doi.org/10.1139/cjfas-2017-0143
2. Thorson, J.T., Jensen, O.P., Zipkin, E.F., 2014. How variable is recruitment for exploited marine fishes? A hierarchical model for testing life history theory. Can. J. Fish. Aquat. Sci. 71, 973–983. https://doi.org/10.1139/cjfas-2013-0645
3. Prince, J., Harford, W.J., Taylor, B.M., Lindfield, S.J., 2022. Standard histological techniques systematically under‐estimate the size fish start spawning. Fish and Fisheries 23, 1507–1516. https://doi.org/10.1111/faf.12702
